\begin{abstract}
  There is currently no rigorous way to reason about the causes of events in a
  real-world distributed system. \jmh{Over-strong. There are formal ways (you're about to say so!), they're just crude. I do think you can find a way to plant a historical flag here that's valid, but you need to be careful.} Causality, from the distributed systems
  literature, can be used to compute the causal history of an arbitrary event
  in a distributed system, but the causal history of an event is a vast
  overapproximation of the true causes. Data provenance, from the database
  literature, describes precisely the cause of a particular output of a
  relational query, but data provenance is limited to the domain of static
  relational databases. In this paper, we present \defword{\watprovenance{}}: a
  novel form of provenance that unifies causality and data provenance. Given an
  arbitrary time-varying state machine, \watprovenance{} describes precisely \jmh{``precisely'' may be a bit strong ... can you prove such a claim?}
  why the state machine produces a particular output when given a particular
  input. This enables system developers to systematically reason about the
  causes of events in real-world distributed systems. We find that
  automatically extracting the \watprovenance{} of an arbitrary state machine
  is often intractable \jmh{``intractable'' has a specific meaning in complexity and I don't think you mean that; pick another term} and sometimes impossible due to their complex
  implementations \jmh{shift in grammar from singular to plural... you may need to rework sentence}. Fortunately, many distributed systems components have simple
  interfaces from which we can directly compute~\jmh{this almost sounds like an automated synthesis approach; be more explicit about human specification} their \watprovenance{} using a
  technique we call \defword{\watprovenance{} specifications}. Using \watprovenance{} as a theoretical foundation, we implement a prototype
  distributed debugging framework called \fluent{} and empirically demonstrate
  that \watprovenance{} makes debugging distributed systems significantly
  easier than what is possible with existing techniques.
\end{abstract}
