\section{\WatProvenance{} Specifications}\seclabel{WatProvSpecs}
Now that we have defined \watprovenance{}, we turn to the matter of computing it.

\subsection{Provenance Specifications}
Automatically computing the \watprovenance{} for an \emph{arbitrary}
distributed system component, which we dub a \defword{black box}, is often
intractable and sometimes impossible. Computing the \watprovenance{} of a black
box requires that we analyze the black box's implementation to extract the
relationships between the inputs and outputs of the black box. Because black
box implementations can be large and complex, this program analysis is almost
always intractable. Worse, we may not have access to the source code of the
black box at all. For example, cloud services like Amazon S3 or Google Cloud
Spanner have proprietary implementations. In this case, automatically computing
\watprovenance{} is impossible.

\newcommand{\kvget}{\texttt{get}}
\newcommand{\kvset}{\texttt{set}}

Though automatically computing the \watprovenance{} for an \emph{arbitrary}
black box is intractable, we can take advantage of the fact that many
real-world black boxes are far from arbitrary. Many black boxes have complex
implementations but are designed with very simple interfaces. This allows us
to sidestep the issue of \emph{inferring} \watprovenance{} from an
implementation and instead \emph{specify} \watprovenance{} directly from an
interface. That is, we can write a \defword{\watprovenance{} specification}: a
function that---given a trace $T$ and request $i$---directly returns the
\watprovenance{} $\Wat{M, T, i}$ for a black box modeled as state machine $M$.

For example, if we restrict our attention to the \kvget{} and \kvset{} API of
Redis, then the \watprovenance{} specification of a \kvget{} request is
trivial: the \watprovenance{} of a \kvget{} request for key $k$
includes only the most recent \kvset{} to $k$. Redis is implemented in over
50,000 lines of C. Analyzing this body of code and inferring the
\watprovenance{} of a \kvget{} request is infeasible using modern program
analysis techniques. \Watprovenance{} specification avoids this issue entirely
and instead specifies the \watprovenance{} in a single line of text.

Moreover, codifying this one-line \watprovenance{} specification is expectedly
straightforward. \Watprovenance{} specifications are implemented as functions
that take in a trace $T$ and an input $i$ and return the \watprovenance{} of
$i$. Thus, \watprovenance{} specifications can be written in any programming
language and can use any language features available.  \Watprovenance{}
specifications do not have to be written using any special domain specific
language or using any restricted subset of a language.

As a simple example, we provide a Python implementation of the one line
\watprovenance{} specification in \figref{RedisProvSpec}.  The specification,
\texttt{get\_prov}, takes in a trace \texttt{T} and a \kvget{} request
\texttt{i} for key $k$. Redis requests are represented as objects of type
\texttt{Request} with subclasses \texttt{GetRequest} and \texttt{SetRequest}.
\texttt{get\_prov} iterates through the trace in reverse order, looking for a
\kvset{} request to key $k$. If such a \kvset{} request is found,
\texttt{get\_prov} returns it. Otherwise, \texttt{get\_prov} returns an empty
witness.
% Recall that the \watprovenance{} of a request is a set of witnesses closed
% under supertrace in $T$, which is why \texttt{get\_prov} returns either a set
% of the singleton witness \texttt{[a]} or a set of the empty witness
% \texttt{[]}.

{\input{figures/redis_prov_spec}}

In \secref{Fluent}, we present a prototype implementation of a system for
writing \watprovenance{} specifications and describe the details of how the
system collects traces. In this section, we omit these details and focus on
the concepts behind \watprovenance{} specifications.

\subsection{Examples}
The \watprovenance{} specification of a Redis \kvget{} request is particularly
simple, but in our experience this simplicity is not completely uncommon. We
now survey a variety of commonly used black boxes that have relatively simple
\watprovenance{} specifications. Later, we discuss some black boxes for which
writing \watprovenance{} specifications is more difficult.

\paragraph{Key-Value Stores}
We have already seen a \watprovenance{} specification for the \kvget{} and
\kvset{} API of Redis. We can easily extend our \watprovenance{} specification
to handle more of Redis' API. For example, consider the operations
\texttt{append}, \texttt{decr}, \texttt{decrby}, \texttt{incr}, and
\texttt{incrby} that all modify the value associated with a particular key.
With these operations present in a trace, the \watprovenance{} specification
for a \kvget{} request to key $k$ now includes the most recent set to $k$ and
all subsequent modifying operations to key $k$.

\paragraph{Object Stores}
We can also write \watprovenance{} specifications for storage systems that are
more complex than key-value stores. For example, consider an object store like
Amazon S3 where users can create, move, copy, list, and get buckets and
objects. The \watprovenance{} specification for the get of an object $o$ in
bucket $b$ includes the most recent creation of the bucket $b$ and the most
recent creation of $o$. If a bucket or object was created by a move, then the
provenance also includes the provenance of the moved bucket or object. The
\watprovenance{} of a request to list the contents of a bucket includes the
most recent creation of the bucket, the most recent creation of every object in
the bucket, and the deletion of any object that was previously in the bucket.

\paragraph{Distributed File Systems}
We can also specify the \watprovenance{} of a distributed file system like NFS
or HDFS. A \watprovenance{} specification of a request to read a byte range
from a file includes the most recent creation of the file, the most recent
creation of the parent directories of the file, and the most recent writes that
overlap with the requested byte range. If the file was created by moving
another file, then the \watprovenance{} specification also includes the
provenance of the file that was moved.

\paragraph{Coordination Services}
Systems use coordination services like Apache
Zookeeper~\cite{hunt2010zookeeper} and Chubby~\cite{burrows2006chubby} for
leader election, mutual exclusion, etc. Take Zookeeper as an example.
Zookeeper's API resembles that of a file system; users can create, delete,
write, and read file-like objects called znodes. Though the implementation of
Zookeeper and HDFS are radically different, their APIs (and thus their
\watprovenance{} specifications) are similar. For example, the \watprovenance{}
specification of a request to read a znode includes the most recent creation of
the znode and the most recent creations of all ancestor znodes.

\paragraph{Load Balancers}
Consider a load balancer, like HAProxy, that is balancing load across a set
$s_1, \ldots, s_n$ of $n$ servers. Periodically, a server $s_i$ sends a
heartbeat to the load balancer that includes $s_i$'s average load for the last
five minutes. Whenever the load balancer receives a message from a client, it
forwards the message to the server $s_i$ that is least loaded. Modelling the
forwarding decision $s_i$ as the output of the load balancer, the
\watprovenance{} specification for the forwarding decision includes the most
recent heartbeat message from the least-loaded server.

\paragraph{Stateless Services}
A stateless service is a service that can be modelled as a state machine with a
single state. Given a request, a stateless service always produces the same
reply, no matter what other requests it has already serviced.  For example, a
web server serving a static website is stateless; it replies to all requests
with the same website. \Watprovenance{} specifications of a stateless service
are trivial. Requests are completely independent, so the \watprovenance{} of
any request consists only of the empty witness.

% \subsection{Discussion}
% These examples show that many popular black boxes have simple \watprovenance{}
% specifications and that many \watprovenance{} specifications are similar to one
% another. We posit that this is not a coincidence. System designers
% design their systems to be as simple and as familiar as possible. If a black
% box had a very complex API with very complex dependencies between requests,
% then it would be challenging for a user to become proficient in using the
% system.
%
% However, not \emph{every} black box has a simple provenance specification. Take
% for example, a relational database where users can insert tuples and issue
% queries (as with \exampleref{WatExampleSetDiff}). As \clmref{WatSubsumesWhy}
% shows, writing a \watprovenance{} specification for a particular query is
% tantamount to specifying the \whyprovenance{} of the query. Worse, if the query
% includes non-monotone constructs like set difference or aggregation that are
% beyond the scope of \whyprovenance{}, the \watprovenance{} specification
% becomes even more challenging to write.
%
% Fortunately, we can mitigate (though not completely eliminate) some of the
% difficulties of writing complex \watprovenance{} specifications using a simple
% yet effective technique we call \defword{exhaustive \watprovenance{} checking}:
% a mechanism for unit testing provenance specifications. Informally, we can
% directly compute the \watprovenance{} for a particular request with respect to
% trace $T$ by exhaustively enumerating every possible subtrace of $T$ and
% checking if it is a witness. Formally, given a state machine $M$ with start
% state $s_0$, a trace $T$, and an input $i$, we can automatically compute the
% \watprovenance{} of $i$ by exhaustively computing $\epsilon^*(s_0, T'i)$ for
% every subtrace $T'$ of $T$. Doing so, we can automatically compute the
% \watprovenance{} of a particular request and verify that it matches the
% \watprovenance{} produced by a provenance specification. Even with exhaustive
% \watprovenance{} checking, complex \watprovenance{} specifications can be
% challenging to write. But, exhaustive \watprovenance{} checking enables a
% developer to find subtle bugs that are otherwise difficult to discover.  Also
% note that exhaustive \watprovenance{} checking runs in superexponential time
% with respect to the length of $T$, so it is effective for small unit tests but
% impractical for computing the \watprovenance{} of real-world traces.
%
% Moreover, we can also take steps to mitigate \emph{redundantly} writing complex
% \watprovenance{} specifications.  The effort of writing a \watprovenance{}
% specification need not be duplicated. Black boxes like Redis are written by a
% handful of developers but used by thousands of others.  Provenance
% specifications are no different. Once a single developer writes a provenance
% specification for a particular black box, all other users of the black box can
% leverage it.

\subsection{Limitations}\seclabel{WatProvSpecsLimitations}
Though \watprovenance{} specifications are often simple to write, they are not
a panacea. Here, we discuss some limitations of \watprovenance{}
specifications.

\paragraph{Complex \watprovenance{} specifications}
There are some black boxes for which writing a \watprovenance{} specification
is inherently very difficult. For example, consider a state machine that
implements an online support vector machine (SVM)~\cite{hearst1998support,
laskov2006incremental, bordes12huller}. Clients can either (a) submit training
data to the state machine to update the model or (b) submit test data to the
state machine for classification. The \watprovenance{} of a classification
request includes only the support vectors of the model at the time of
classification. Writing a \watprovenance{} specification to identify these
support vectors is possible but very difficult. Such a \watprovenance{}
specification would likely be as complicated as the state machine itself.

% However, not \emph{every} black box has a simple provenance specification. Take
% for example, a relational database where users can insert tuples and issue
% queries (as with \exampleref{WatExampleSetDiff}). As \clmref{WatSubsumesWhy}
% shows, writing a \watprovenance{} specification for a particular query is
% tantamount to specifying the \whyprovenance{} of the query. Worse, if the query
% includes non-monotone constructs like set difference or aggregation that are
% beyond the scope of \whyprovenance{}, the \watprovenance{} specification
% becomes even more challenging to write.

% \paragraph{Nondeterministic Black Boxes}
% The definition of \watprovenance{} assumes a \emph{deterministic} state
% machine, yet some black boxes are nondeterministic. For example, the behavior
% of many weakly consistent distributed systems depend on the non-deterministic
% ordering of messages in the network. \Watprovenance{} specifications cannot be
% applied to nondeterministic distributed systems like this. We leave a
% generalization of \watprovenance{} and \watprovenance{} specifications to
% nondeterministic state machines as an interesting avenue for future work.

\paragraph{Buggy black boxes}
We have thus far tacitly assumed that black boxes like Redis faithfully
implement their advertised interfaces. However, if a black box is buggy and
deviates from its expected behavior, then a \watprovenance{} specification will
produce erroneous provenance. In other words, \watprovenance{} specifications
are written with respect to the semantics of an abstract state machine. If a
particular black box does not concretely implement these semantics, then the
\watprovenance{} specification is incorrect.

% Moreover, if we have confidence in the correctness of a \watprovenance{}
% specification, then we can use exhaustive \watprovenance{} checking on small
% inputs to identify potential bugs in a black box.  If the \watprovenance{}
% computed by an exhaustive \watprovenance{} check differs from the
% \watprovenance{} returned by a specification, it's an indication that the black
% box might be deviating from its intended specification. Note that if we are
% confident that the black box implementation is correct, then we can use
% exhaustive \watprovenance{} checking to find bugs in a \watprovenance{}
% specification, and if if we have confidence that the \watprovenance{}
% specification is correct, then we can use exhaustive \watprovenance{} checking
% to find bugs in the black box implementation.
